---
title: "Gumbel hd32"
output:
  bookdown::html_document2:
    toc: false
    self_contained: true
---

\vspace{-0.1in}

Install and load required packages
```{r, message=FALSE, warning=FALSE}
R_packages <- c("bookdown", "copula", "ggplot2", "gridExtra", "ks", "plot3D", "viridis",
                "ggpubr", "grid", "cowplot")
options(repos = c(CRAN="http://cran.rstudio.com"))
if (!requireNamespace("librarian", quietly = TRUE)) install.packages("librarian")
librarian::shelf(R_packages)
```

Set parameters
```{r}
a <- 4 # copula parameter
gsz <- 100 # grid size
dim <- 3 # dimensions of copula
rname <- "Gumbel-hd32"
```

# Gumbel Copula

Define the Gumbel copula 
```{r}
gumbel_cop <- gumbelCopula(param = a, dim = dim)
```

Create grid on the hypercube
```{r}
# sequence of points in each dimension
seqs <- replicate(dim, seq(0.01, 0.99, length.out = gsz), simplify = FALSE)

# create names for the dimensions
names(seqs) <- paste0("u", 1:dim)

# construct the grid
grid <- expand.grid(seqs)
```

Compute copula density on $[0,1]^2$ copula grid
```{r}
copula_dens <- dCopula(as.matrix(grid), gumbel_cop)
```

For graphs: join grid and copula density
```{r}
gr_dens <- cbind(grid, round(copula_dens,4))
colnames(gr_dens)[ncol(gr_dens)] <- "copula_dens"
```

## Plots over x-y subgrid
```{r}
ux <- "u1"
uy <- "u3"

free_dims <- c(ux, uy)
all_dims  <- grep("^u[0-9]+$", names(gr_dens), value = TRUE)
fixed_dims <- setdiff(all_dims, free_dims)

# compute target values (averages of fixed dims)
target_vals <- colMeans(gr_dens[, fixed_dims, drop = FALSE])

# find closest grid values to those averages
closest <- vapply(fixed_dims, function(dim) {
  vals <- unique(gr_dens[[dim]])
  vals[which.min(abs(vals - target_vals[dim]))]
}, numeric(1))

# filter grid at fixed values
slice_df <- gr_dens
for(dm in fixed_dims){
  slice_df <- slice_df[slice_df[[dm]] == closest[[dm]], ]
}

# dataframe for plotting
plot_df <- slice_df[, c(ux, uy, "copula_dens")]
```

Plot copula density on $[0,1]^2$ copula grid 3D
```{r, fig.cap = "&nbsp;", fig.width=6, fig.height=4}
z_mat_3d <- matrix(as.numeric(plot_df[,"copula_dens"]), nrow = gsz, byrow = FALSE)
z_mat_3d <- pmin(z_mat_3d, 10) # Truncate for better visualization

persp3D(
  x = seqs[[ux]],
  y = seqs[[uy]],
  z = z_mat_3d,
  theta = 25,
  phi = 50,
  col = viridis::viridis(50),
  xlab = "u1", ylab = "u2", zlab = "density",
  main = "Bivariate Gumbel Copula Density (a = 2)",
  cex.main = 0.6, cex.lab = 0.6,
  colkey = list(length = 0.6, width = 0.6, cex.axis = 0.6)
) 
```

Plot copula density on $[0,1]^2$ copula grid contour
```{r, fig.cap = "&nbsp;", fig.width=6, fig.height=4}
z_mat_2d <- matrix(as.numeric(plot_df[,"copula_dens"]), nrow = gsz, byrow = FALSE)
z_mat_2d <- pmin(z_mat_2d, 8) # Truncate for better visualization

filled.contour(
  x = seqs[[ux]], y = seqs[[uy]], z = z_mat_2d,
  color.palette = viridis::viridis,
  xlab = "", ylab = "",
  plot.axes = {axis(1, at = seq(0.2, 0.8, by = 0.2), line = -0.05); axis(2, line = -3.5)},  
  plot.title = title(main = "Copula Density: Contour Plot", cex.main = 0.8),
  key.title = title(main = "Density", cex.main = 0.6), asp = 1, frame.plot = FALSE,
  colkey = list(addlines = FALSE, length = 0.6, width = 0.5, cex.clab = 0.8)
)
```    

# Simulated Data

Simulate data
```{r}
n <- 1000 # simulated sample size
set.seed(123456) 
cs <- rCopula(n, gumbel_cop)  
```

Plot function 
```{r}
plot_lower_triangular <- function(cs_df, title = NULL,
                                  point_alpha = 0.4, point_size = 0.2) {
  d <- ncol(cs_df)
  vars <- colnames(cs_df)
  if (d < 2) stop("Need at least 2 columns for pairwise plots.")
  
  my_scatter <- function(data, xvar, yvar) {
    ggplot(data, aes(x = {{xvar}}, y = {{yvar}})) +
      geom_point(alpha = point_alpha, size = point_size) +
      coord_fixed(ratio = 1) +
      labs(x = "", y = "") +
      theme_minimal() +
      theme(
        plot.margin  = unit(c(0,0,0,0), "in"),
        panel.grid   = element_blank(),
        axis.text    = element_blank(),
        axis.ticks   = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5)
      )
  }
  
  empty_panel <- ggplot() + theme_void()
  
  # build full (d-1) x (d-1) matrix
  plot_list <- vector("list", (d-1)*(d-1))
  
  for (i in 2:d) {
    for (j in 1:(d-1)) {
      idx <- (i-2)*(d-1) + j
      if (j < i) {
        plot_list[[idx]] <- my_scatter(cs_df, .data[[vars[j]]], .data[[vars[i]]])
      } else {
        plot_list[[idx]] <- empty_panel
      }
    }
  }
  
  combined_plot <- ggarrange(
    plotlist = plot_list,
    ncol = d-1,
    nrow = d-1,
    align = "hv"
  )
  
  if (!is.null(title)) {
    combined_plot <- annotate_figure(
      combined_plot,
      top = text_grob(title, size = 8)
    )
  }
  
  return(combined_plot)
}
```

Scatter plot of simulated draws from copula, pairwise dimensions
```{r, fig.width=4, fig.height=4, fig.cap = "&nbsp;"}
cs_df <- as.data.frame(cs)
colnames(cs_df) <- paste0("U", 1:dim)
csplot <- plot_lower_triangular(cs_df, title="Gumbel 2") 
csplot <- ggdraw(csplot) +
          theme(plot.background = element_rect(color = "gray", fill = NA, size = 0.5),
                plot.margin = unit(c(0., 0.1, -0.1, -0.1), "in"))

plot_file = paste0("../graphs_sim/",rname,"_true.png")
ggsave(plot_file, plot = csplot, width = 4, height = 4, units = "in", dpi = 300)

csplot
```

# Sparse Bernstein Copula

## Set parameters
```{r}
true_dens <- copula_dens
sample_sizes <- c(100, 200, 300, 400, 500, 1000, 10000)
#sample_sizes <- c(100, 200, 300, 400, 500, 1000)
kmax <- 10 # number of batches for each sample size
```

Save true density
```{r}
write.table(true_dens, file = paste0("../data_sim/",rname,"_dat_cdens_true.txt"),
            sep = ",", row.names = FALSE, col.names = FALSE)
```

## Generate Data of Varying Sample Sizes for Fortran Estimation
```{r}
set.seed(123456)

for (n in sample_sizes) {

  for (k in 1:kmax) {
  
    cs <- rCopula(n, gumbel_cop) 
    
    # Write out for Fortran input
    write.table(cs, file = paste0("../data_sim/",rname,"_dat_",n,"_",k,".csv"),
                sep = ",", row.names = FALSE, col.names = FALSE)
  }
}
```

Run Fortran estimation of SBP copula

## Plot Simulated Draws from SBP copula
```{r, fig.width=4, fig.height=4, fig.cap = "&nbsp;"}
n  <- 1000 # for sample size
k <- 1 # for data from batch k
csim_SBP <- read.table(file = paste0("../output_sim/",rname,"_csim_",n,"_",k,".out"), header = FALSE)
csim_SBP_df <- as.data.frame(csim_SBP)
colnames(csim_SBP_df) <- paste0("U", seq_len(ncol(csim_SBP_df)))

csplot_sim <- plot_lower_triangular(csim_SBP_df, title="SBP copula") 
csplot_sim <- ggdraw(csplot_sim) +
                theme(plot.background = element_rect(color = "gray", fill = NA, size = 0.5),
                plot.margin = unit(c(0., 0.1, -0.1, -0.1), "in"))

plot_file_sim = paste0("../graphs_sim/",rname,"_SBP.png")
ggsave(plot_file_sim, plot = csplot_sim, width = 4, height = 4, units = "in", dpi = 300)

csplot_sim
```

## Evaluate Kullback-Leibler Distance from True Density
```{r}
KL <- numeric() 
KLse <- numeric()

for (n in sample_sizes) {

  KL_k <- numeric() 
  epsilon <- 1e-10 # Add small epsilon to avoid division by 0
  
  for (k in 1:kmax) {
    cdens_SBP <- read.table(file = paste0("../output_sim/",rname,"_cdens_",n,"_",k,".out"), header = FALSE)
    cdens_SBP <- as.numeric(unlist(cdens_SBP))
    ratio <- (true_dens + epsilon)/(cdens_SBP + epsilon)
    KL_k_val <- log(ratio)*true_dens # Compute KL integrand
    KL_km <- mean(KL_k_val) # approximate integral over the grid
    KL_k <- c(KL_k, KL_km) # add new row
  }

  KL <- c(KL, mean(KL_k)) # add new row
  KLse <- c(KLse, sd(KL_k)/sqrt(length(KL_k))) # add new row
  
}

print(round(KL, 2))
print(round(KLse, 2))
```
